---
title: "Breathing Together Project: G2-4"
subtitle: "Host Transcriptomics"
author: "Matthew Macowan"
date: "`r format(Sys.time())`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 4
---

Copyright (c) 2023, Mucosal Immunology Lab, Monash University, Melbourne, Australia.

# Environment setup

## 1) Load packages

```{r load packages, eval=TRUE, collapse=TRUE, echo=FALSE}
notebook.name <- 'BT_host_transcriptomics_analysis'

# Get R version and OS information
R.version$version.string
R.version$platform

# Load R packages
pkgs <- c('knitr', 'here', 'SummarizedExperiment', 'biomaRt', 'edgeR', 'limma')

pacman::p_load(char = pkgs)

for (pkg in pkgs) {
  print(paste0(pkg, ': ', packageVersion(pkg)))
}

# Set seed
set.seed(2)
```

## 2) Process data and prepare matrix for multi-omics

The RNA sequencing raw fastq files have been processed using the [NF-CORE rnaseq pipeline](https://nf-co.re/rnaseq).
Briefly, reads undergo initial QC with FastQC, UMIs are extracted with UMI-tools, and reads undergo adapter removal and quality trimming using Trim Galore!
Next, genomic contaminants are removed via BBSplit, and ribosomal components are removed via SortMeRNA.
Reads are aligned and quantified via a combination of STAR and Salmon tools, and sorted and indexed using SAMtools and dereplicated via UMI-tools.
Outputs are provided in `.rds` file format as SummarizedExperiment objects, with bias-corrected gene counts without an offset (`salmon.merged.gene_counts_length_scaled.rds`).

There are two matrices provided to us: counts and abundance.

- The abundance matrix is the scaled and normalised transcripts per million (TPM) abundance. TPM explicitly erases information about library size. That is, it estimates the relative abundance of each transcript proportional to the total population of transcripts sampled in the experiment. Thus, you can imagine TPM, in a way, as a partition of unity — we want to assign a fraction of the total expression (whatever that may be) to transcript, regardless of whether our library is 10M fragments or 100M fragments.
- The counts matrix is a re-estimated counts table that aims to provide count-level data to be compatible with downstream tools such as DESeq2.
- The `tximport` package has a single function for importing transcript-level estimates. The type argument is used to specify what software was used for estimation. A simple list with matrices, `"abundance"`, `"counts"`, and `"length"`, is returned, where the transcript level information is summarized to the gene-level. Typically, abundance is provided by the quantification tools as TPM (transcripts-per-million), while the counts are estimated counts (possibly fractional), and the `"length"` matrix contains the effective gene lengths. The `"length"` matrix can be used to generate an offset matrix for downstream gene-level differential analysis of count matrices.

As such, there are a few things we need to keep in mind:

- For ordination, we should use the TPM data.
- For limma DE analysis, because limma does not use an offset matrix store in `y$offset`, it is recommended to use the scaled counts generated from abundances, either `"scaledTPM"` or `"lengthScaledTPM"`.
- For edgeR, you need to generate counts from abundances.

```{r}
# Import the bias-corrected counts from STAR Salmon
rna_bronch <- readRDS(here('input', '03_HostTranscriptomics', 'StarSalmon_BronchBrush', 'salmon.merged.gene_counts_length_scaled.rds'))

# Get Ensembl annotations
ensembl <- useMart('ensembl', dataset = 'hsapiens_gene_ensembl')

ensemblIDsBronch <- rownames(rna_bronch)

gene_list <- getBM(attributes = c('ensembl_gene_id', 'hgnc_symbol', 'gene_biotype'),
                   filters = 'ensembl_gene_id', values = ensemblIDsBronch, mart = ensembl)
colnames(gene_list) <- c("gene_id", "hgnc_symbol", "gene_biotype")
gene_list <- filter(gene_list, !duplicated(gene_id))

# Ensure that only genes in the STAR Salmon outputs are kept for the gene list
rna_bronch <- rna_bronch[rownames(rna_bronch) %in% gene_list$gene_id, ]

# Add the ENSEMBL data to the rowData element
rowData(rna_bronch) <- merge(gene_list, rowData(rna_bronch), by = "gene_id", all = FALSE)

# Load the RNA metadata
metadata_rna <- readRDS(here('input', '00_Metadata', 'metadata_rna_clean.rds'))

# Sort the metadata rows to match the order of the abundance data
metadata_bronch <- metadata_rna
rownames(metadata_bronch) <- metadata_bronch$Bronch_RNA_barcode
metadata_bronch <- metadata_bronch[colnames(rna_bronch),]

# Create a DGEList from the SummarizedExperiment object
rna_bronch_dge <- DGEList(assay(rna_bronch, 'counts'), 
                          samples = metadata_bronch, 
                          group = metadata_bronch$group,
                          genes = rowData(rna_bronch),
                          remove.zeros = TRUE)

# Filter the DGEList based on the BT group information
design_bronch <- model.matrix(~ group, data = rna_bronch_dge$samples)
keep_bronch_min100 <- filterByExpr(rna_bronch_dge, design_bronch, min.count = 100)
rna_bronch_dge_min100 <- rna_bronch_dge[keep_bronch_min100, ]

# Calculate norm factors and perform voom normalisation
rna_bronch_dge_min100 <- calcNormFactors(rna_bronch_dge_min100)
rna_bronch_dge_min100 <- voom(rna_bronch_dge_min100, design_bronch, plot = TRUE)

# Add the normalised abundance data from STAR Salmon and filter to match the counts data
rna_bronch_dge_min100$abundance <- as.matrix(assay(rna_bronch, 'abundance'))[keep_bronch_min100, ]

# Select protein coding defined genes only
rna_bronch_dge_min100 <- rna_bronch_dge_min100[rna_bronch_dge_min100$genes$gene_biotype == "protein_coding" & rna_bronch_dge_min100$genes$hgnc_symbol != "", ]

# Add symbol as rowname
rownames(rna_bronch_dge_min100) <- rna_bronch_dge_min100$genes$gene_name

# Save the DGEList
saveRDS(rna_bronch_dge_min100, here('input', '03_HostTranscriptomics', 'rna_bronch_dge_min100.rds'))

# Prepare the voom-corrected gene expression data for MOFA+
# We need to add an 'S' to the beginning of the samples as they shouldn't start with numeric values
host_rna_bronch_matrix_min100 <- rna_bronch_dge_min100$E
colnames(host_rna_bronch_matrix_min100) <- paste0('S', rna_bronch_dge_min100$targets$Subject)
host_rna_bronch_matrix_min100 <- host_rna_bronch_matrix_min100[, !colnames(host_rna_bronch_matrix_min100) %in% outliers]

# Save the matrix for multi-omics
saveRDS(host_rna_bronch_matrix_min100, here('input', '06_MOFA', 'host_rna_bronch_matrix_min100.rds'))
```