---
title: "Breathing Together Project - Groups 2-4"
subtitle: "Small Molecules (Metabolomics and Lipidomics)"
author: "Matthew Macowan"
date: "`r format(Sys.time())`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 4
---

Copyright (c) 2023, Mucosal Immunology Lab, Monash University, Melbourne, Australia.

# Environment setup

## 1) Load packages

```{r load packages, eval=TRUE, collapse=TRUE, echo=FALSE}
notebook.name <- '04_SmallMolecules'

# Get R version and OS information
version$version.string
version$platform

#devtools::install_github("yogevherz/plotme")

# Load R packages
pkgs <- c('knitr', 'markdown', 'tidyverse', 'dplyr', 'ggplot2', 'here', 'data.table',
          'kableExtra', 'pmp', 'SummarizedExperiment', 'S4Vectors')

pacman::p_load(char = pkgs)

for (pkg in pkgs) {
  print(paste0(pkg, ': ', packageVersion(pkg)))
}

# Set seed
set.seed(2)

# Set BT group colours
bt_cols <- c('2_Preschool_control' = 'darkseagreen3', 
             '3_Preschool_wheeze' = 'orange3',
             '4_School_asthma' = 'indianred')

# Load custom scripts
source(here::here('scripts', 'rotate_df.R'))
source(here::here('scripts', 'plot_list_to_pdf.R'))
source(here::here('scripts', 'add_hmdb.R'))
source(here::here('scripts', 'add_lmsd.R'))
source(here::here('scripts', 'keep_annotated.R'))
source(here::here('scripts', 'save_curation_table.R'))
```

# Import Data

## 1) Metadata

```{r}
# Load the sample metadata (and set the column data types)
smallmol_metadata <- read_csv(here('input', '00_Metadata', 'smallmol_metadata.csv'), 
                                  col_types = 'cncfnnffccnfffffnnfcccccffffffffffffffnfnnnnnnnnnf')
# Save to RDS file
saveRDS(smallmol_metadata, here('input', '00_Metadata', 'smallmol_metadata.rds'))
```

## 2) MS-DIAL outputs

We will load in both the positive and negative ionisation mode height matrices from the MS-DIAL pipeline. We need to remove the MS/MS columns, as these were acquired in a different manner to the other columns.

Then, we can rbind the two tables together after denoting the original ionisation mode in the row names. Because the QC data is incorporated within each feature (row), pmp can pre-process all of our data at once, and normalise the entire dataset.

```{r}
# Set the base MS-DIAL output folder location
msdial_path <- here('input', '04_SmallMolecules', '02_LCMS_data_curation')
```

### i) BAL metabolomics data

```{r}
# Load metabolomics height data
metab_bal_pos <- read_csv(here(msdial_path, 'BAL_metabolomics', 'BT_BAL_metab_positive_height.csv'), skip = 4)
metab_bal_neg <- read_csv(here(msdial_path, 'BAL_metabolomics', 'BT_BAL_metab_negative_height.csv'), skip = 4)

# Remove the MS2 samples (not acquired in the same manner)
metab_bal_pos <- metab_bal_pos[, !(names(metab_bal_pos) %in% c('MSMS_pos', 'MSMS_neg'))]
metab_bal_neg <- metab_bal_neg[, !(names(metab_bal_neg) %in% c('MSMS_pos', 'MSMS_neg'))]

# Separate into intensity and information data.frames for the SummarizedExperiment object
metab_bal_pos_counts <- as.matrix(metab_bal_pos[, c(36:61, 63:69, 71:85)])
metab_bal_neg_counts <- as.matrix(metab_bal_neg[, c(36:61, 63:69, 71:85)])

metab_bal_pos_info <- metab_bal_pos[, 1:35]
metab_bal_neg_info <- metab_bal_neg[, 1:35]

# Rename the data to indicate ionisation mode (using the MS-DIAL alignment ID)
metab_bal_pos_rownames <- paste0(metab_bal_pos_info$`Alignment ID`, '_met_pos')
metab_bal_neg_rownames <- paste0(metab_bal_neg_info$`Alignment ID`, '_met_neg')

rownames(metab_bal_pos_counts) <- metab_bal_pos_rownames
rownames(metab_bal_pos_info) <- metab_bal_pos_rownames
rownames(metab_bal_neg_counts) <- metab_bal_neg_rownames
rownames(metab_bal_neg_info) <- metab_bal_neg_rownames

# Merge the postive and negative ionisation modes using rbind
metab_bal_counts <- rbind(metab_bal_pos_counts, metab_bal_neg_counts)
metab_bal_info <- rbind(metab_bal_pos_info, metab_bal_neg_info)

# Remove extraneous files
rm(list = c('metab_bal_pos', 'metab_bal_neg', 'metab_bal_pos_counts', 'metab_bal_neg_counts',
            'metab_bal_pos_info', 'metab_bal_neg_info', 'metab_bal_pos_rownames',
            'metab_bal_neg_rownames'))
```

### ii) BAL lipidomics data

```{r}
# Load metabolomics height data
lipid_bal_pos <- read_csv(here(msdial_path, 'BAL_lipidomics', 'BT_BAL_lipid_positive_height.csv'), skip = 4)
lipid_bal_neg <- read_csv(here(msdial_path, 'BAL_lipidomics', 'BT_BAL_lipid_negative_height.csv'), skip = 4)

# Remove the MS2 samples (not acquired in the same manner)
lipid_bal_pos <- lipid_bal_pos[, !(names(lipid_bal_pos) %in% c('MSMS_pos', 'MSMS_neg'))]
lipid_bal_neg <- lipid_bal_neg[, !(names(lipid_bal_neg) %in% c('MSMS_pos', 'MSMS_neg'))]

# Separate into intensity and information data.frames for the SummarizedExperiment object
lipid_bal_pos_counts <- as.matrix(lipid_bal_pos[, c(36:61, 63:69, 71:85)])
lipid_bal_neg_counts <- as.matrix(lipid_bal_neg[, c(36:61, 63:69, 71:85)])

lipid_bal_pos_info <- lipid_bal_pos[, 1:35]
lipid_bal_neg_info <- lipid_bal_neg[, 1:35]

# Rename the data to indicate ionisation mode (using the MS-DIAL alignment ID)
lipid_bal_pos_rownames <- paste0(lipid_bal_pos_info$`Alignment ID`, '_lip_pos')
lipid_bal_neg_rownames <- paste0(lipid_bal_neg_info$`Alignment ID`, '_lip_neg')

rownames(lipid_bal_pos_counts) <- lipid_bal_pos_rownames
rownames(lipid_bal_pos_info) <- lipid_bal_pos_rownames
rownames(lipid_bal_neg_counts) <- lipid_bal_neg_rownames
rownames(lipid_bal_neg_info) <- lipid_bal_neg_rownames

# Merge the postive and negative ionisation modes using rbind
lipid_bal_counts <- rbind(lipid_bal_pos_counts, lipid_bal_neg_counts)
lipid_bal_info <- rbind(lipid_bal_pos_info, lipid_bal_neg_info)

# Remove extraneous files
rm(list = c('lipid_bal_pos', 'lipid_bal_neg', 'lipid_bal_pos_counts', 'lipid_bal_neg_counts',
            'lipid_bal_pos_info', 'lipid_bal_neg_info', 'lipid_bal_pos_rownames',
            'lipid_bal_neg_rownames'))
```

### iii) Merge metabolomics and lipidomics data together

Just as we merged the positive and negative ionisation modes for the metabolomics and lipidomics data individually, we will now merge both into a single small molecule dataset for combined normalisation. 

```{r}
# Check that the column names match
identical(colnames(metab_bal_counts), colnames(lipid_bal_counts))

# Merge the metabolomics and lipidomics data
smallmol_counts <- rbind(metab_bal_counts, lipid_bal_counts)
smallmol_counts <- data.frame(smallmol_counts)  %>%
  mutate(across(.cols = everything(), .fns = as.numeric)) %>%
  as.matrix()
smallmol_info <- rbind(metab_bal_info, lipid_bal_info)

# Remove extraneous files
rm(list = c('metab_bal_counts', 'metab_bal_info', 'lipid_bal_counts', 'lipid_bal_info'))
```

### iv) Import data into SummarizedExperiment

Now that we have our intensity and feature information datasets, before we can import them into a SummarizedExperiment object, we need to define class and group vectors so that pmp knows whether our variables are samples, QCs, or blanks.

The easiest way to achieve this is by getting the first two characters (a substring) of our column names, and using these as indicators of the classes: 'Bl' = blank, 'QC' = QC etc.

```{r}
# Retrieve metadata
smallmol_metadata <- readRDS(here('input', '00_Metadata', 'smallmol_metadata.rds'))

# Create class and group vectors
smallmol_class <- substr(colnames(smallmol_counts), start = 1, stop = 2)
smallmol_group <- substr(colnames(smallmol_counts), start = 1, stop = 2)

# Reorder the metadata rows so that they match the column order of samples in the counts data.frame (ignoring blanks and QCs)
rownames(smallmol_metadata) <- smallmol_metadata$LCMS_ID
smallmol_metadata <- smallmol_metadata[colnames(smallmol_counts),]
rownames(smallmol_metadata) <- smallmol_metadata$LCMS_ID

# Check that the metadata matches the samples
identical(rownames(smallmol_metadata), colnames(smallmol_counts))

# Create the SummarizedExperiment (SE) object
smallmol_SE <- SummarizedExperiment(assays = list(counts = smallmol_counts),
                                    metadata = list(smallmol_metadata),
                                    rowData = list(info = smallmol_info),
                                    colData = DataFrame(class = smallmol_class))
```

# Data pre-processing

## 1) Filtering and normalisation

### i) Filtering

The first filtering steps we will carry out are to replace any so-called "missing" values (i.e. 0 values) with NA so they will be compatible with downstream filtering steps.

After this, we can filter peaks based on the intensity values of our blanks.

```{r}
# Check the original number of features
(dim0 <- dim(smallmol_SE))

# Replace missing values with NA to be compatible with downstream filtering
assay(smallmol_SE) <- replace(assay(smallmol_SE), assay(smallmol_SE) == 0, NA)

# Filter peaks and optionally samples based on blanks
smallmol_filt <- filter_peaks_by_blank(df = smallmol_SE,
                                       fold_change = 5, # 5-fold change
                                       classes = smallmol_SE$class,
                                       remove_samples = TRUE,
                                       remove_peaks = TRUE,
                                       blank_label = 'Bl') # from the class vector
                                           
# Check the number of features/samples remaining
(dim1 <- dim(smallmol_filt))
```

Next, we can perform a few filtering steps based on missing values and degree of variation in the QC samples.

* filter_samples_by_mv: removal of samples containing a user-defined maximum percentage of missing values see documentation.
* filter_peaks_by_fraction: removal of peaks based upon the relative proportion of samples containing non-missing values see documentation.
* filter_peaks_by_rsd: removal of peaks based upon relative standard deviation of intensity values for a given feature within specified QC samples see documentation.

```{r}
# Filter samples based on the percentage of missing values
smallmol_filt <- filter_samples_by_mv(df = smallmol_filt,
                                      max_perc_mv = 0.8)
                                         
# Check the number of features/samples
(dim2 <- dim(smallmol_filt))

# Filter peaks based on missing values across all samples
smallmol_filt <- filter_peaks_by_fraction(df = smallmol_filt,
                                          min_frac = 0.8,
                                          classes = smallmol_filt$class,
                                          method = 'across')

# Check the number of features/samples
(dim3 <- dim(smallmol_filt))

# Filter peaks based on the percentage of variation in the QC samples
smallmol_filt <- filter_peaks_by_rsd(df = smallmol_filt,
                                     max_rsd = 25,
                                     classes = smallmol_filt$class,
                                     qc_label = 'QC')

# Check the number of features/samples
(dim4 <- dim(smallmol_filt))
```

### ii) PQN normalisation and glog scaling

```{r}
# PQN data normalisation
smallmol_norm <- pqn_normalisation(df = smallmol_filt,
                                   classes = smallmol_filt$class,
                                   qc_label = 'QC')
                                      
# Missing values imputation
smallmol_imp <- mv_imputation(df = smallmol_norm,
                              rowmax = 0.7, # max % of missing data allowed in any row
                              colmax = 0.7, # max % of missing data allowed in any column
                              method = 'rf') # or rf, bcpa, sv, mn, md.
                                 
# Data scaling
smallmol_glog <- glog_transformation(df = smallmol_imp,
                                     classes = smallmol_imp$class,
                                     qc_label = 'QC')

opt_lambda_stool <- processing_history(smallmol_glog)$glog_transformation$lambda_opt

glog_plot_optimised_lambda(df = smallmol_imp,
                           optimised_lambda = opt_lambda_stool,
                           classes = smallmol_imp$class,
                           qc_label = 'QC')

# Save file
saveRDS(smallmol_glog, here(msdial_path, 'BAL_smallmol', 'smallmol_glog.rds'))
```

### iii) PCoA

```{r}
# Perform the PCA and retrieve the explained variance values
PCA <- prcomp(t(assay(smallmol_glog)), center = TRUE)
varexp <- c(summary(PCA)$importance[2,1]*100,
            summary(PCA)$importance[2,2]*100)

# Create a dataset for plotting
data_PCA <- cbind(data.frame(Samples = rownames(PCA$x),
                             PC1 = PCA$x[,1],
                             PC2 = PCA$x[,2]),
                  class = smallmol_glog$class)

# Plot results
(PCA_plot <- ggplot(data_PCA, aes(x = PC1, y = PC2)) +
    geom_point(aes(fill = class, color = factor(class))) +
    stat_ellipse(aes(fill = class), geom = 'polygon', type = 't', level = 0.9, alpha = 0.2) +
    labs(title = 'BAL Metabolomics and Lipidomics',
         x = paste0('PC1 ', round(varexp[1], 2), '%'),
         y = paste0('PC2 ', round(varexp[2], 2), '%'))
)
```

## 2) HMDB database

```{r}
# Recover the small molecule glog data
msdial_path <- here('input', '04_SmallMolecules', '02_LCMS_data_curation')
smallmol_glog <- readRDS(here(msdial_path, 'BAL_smallmol', 'smallmol_glog.rds'))

# Load HMDB database
hmdb_df <- readRDS(here('assets', 'hmdb_detected_quantified_v4_20210701.rds'))

# Search annotations in HMDB and add to the SE object
smallmol_glog <- add_hmdb(metab_SE = smallmol_glog,
                          hmdb = hmdb_df,
                          mass_tol = 0.002)
```

## 3) LIPIDMAPS database

```{r}
# Load LMSD database
lmsd_df <- readRDS(here('assets', 'LMSD_2022_02_16.rds'))

# Search LMSD annotations
lmsd_ann_list <- add_lmsd(metab_SE = smallmol_glog,
                          lmsd = lmsd_df,
                          mass_tol = 0.002,
                          cores = 6)

# Confirm rownames match
identical(rownames(smallmol_glog), rownames(lmsd_ann_list$agg_lmsd_df))

# Add LMSD annotations to the SE object
lmsd_agg <- lmsd_ann_list$agg_lmsd_df
rowData(smallmol_glog)$LMSD_NAME <- lmsd_agg$LMSD_NAME
rowData(smallmol_glog)$LMSD_SYSTEMATIC_NAME <- lmsd_agg$LMSD_SYSTEMATIC_NAME
rowData(smallmol_glog)$LMSD_ABBREVIATION <- lmsd_agg$LMSD_ABBREVIATION
rowData(smallmol_glog)$LMSD_CATEGORY <- lmsd_agg$LMSD_CATEGORY
rowData(smallmol_glog)$LMSD_MAIN_CLASS <- lmsd_agg$LMSD_MAIN_CLASS
rowData(smallmol_glog)$LMSD_SYNONYMS <- lmsd_agg$LMSD_SYNONYMS
```

## 4) Generate a consensus (shortname) column

```{r}
# Run keep_annotated function
smallmol_anno <- keep_annotated(metab_SE = smallmol_glog)

# Save SummarizedExperiment
saveRDS(smallmol_anno, here(msdial_path, 'BAL_smallmol', 'smallmol_glog_annotated_only.rds'))
```

## 5) Manual curation

### i) Save curation table

```{r}
# Recover annotated SummarizedExperiment
msdial_path <- here('input', '04_SmallMolecules', '02_LCMS_data_curation')
smallmol_anno <- readRDS(here(msdial_path, 'BAL_smallmol', 'smallmol_glog_annotated_only.rds'))

# Save curation table
save_curation_table(smallmol_anno, here(msdial_path, 'BAL_smallmol', 'curation_table.csv'))
```

### ii) Retrieve curated table and filter

Now that we have gone through and manually curated each of the peaks, we will keep only those with good quality peaks.

```{r}
# Recover annotated SummarizedExperiment
msdial_path <- here('input', '04_SmallMolecules', '02_LCMS_data_curation')
smallmol_anno <- readRDS(here(msdial_path, 'BAL_smallmol', 'smallmol_glog_annotated_only.rds'))

# Read in the curated table
curated_table <- read_csv(here(msdial_path, 'BAL_smallmol', 'curated_table.csv'))

# Filter the annotated SE for quality peaks
smallmol_anno_curated <- smallmol_anno[curated_table$quality_peak,]

# Load LipidLynxX converter names
lipidlynx_anno <- read_csv(here(msdial_path, 'BAL_smallmol', 'LipidLynxX_converted.csv'))[,-1] %>%
  distinct()

# Add LipidLynxX converted names to the SE
lipidlynx <- curated_table %>%
  filter(quality_peak == 'TRUE') %>%
  left_join(lipidlynx_anno, by = c('updated_name' = 'Feature')) %>%
  write_csv(here(msdial_path, 'BAL_smallmol', 'renaming_table.csv'))

# Retrieve renamed data
renamed_table <- read_csv(here(msdial_path, 'BAL_smallmol', 'renamed_table.csv'), show_col_types = FALSE) %>%
  dplyr::select(Ionisation, updated_name, best_peak)
renamed_ordered <- lipidlynx %>%
  left_join(renamed_table, by = c('Ionisation'))

# Add data to the SE
rowData(smallmol_anno_curated)$molecule_name <- renamed_ordered$updated_name.y
rowData(smallmol_anno_curated)$best_peak <- renamed_ordered$best_peak

# Filter for best peak only
smallmol_anno_curated <- smallmol_anno_curated[rowData(smallmol_anno_curated)$best_peak, ]
rowData(smallmol_anno_curated)$best_peak <- NULL

# Remove QC samples
BAL_samples <- str_detect(colnames(smallmol_anno_curated), 'BAL')
smallmol_anno_curated <- smallmol_anno_curated[, BAL_samples]

# Correct sample names to be the Subject ID
smallmol_metadata <- readRDS(here('input', '00_Metadata', 'smallmol_metadata.rds')) %>%
  column_to_rownames(var = 'LCMS_ID')
sample_names <- smallmol_metadata[colnames(smallmol_anno_curated), ]
colnames(smallmol_anno_curated) <- sample_names$Subject_ID

# Save SummarizedExperiment
saveRDS(smallmol_anno_curated, here('input', '04_SmallMolecules', 'smallmol_glog_annotated_curated.rds'))
```

# Data preparation for MOFA integration

## a) Prepare MOFA counts matrix

```{r}
# Retrieve processed dataset
smallmol_SE <- readRDS(here('input', '04_SmallMolecules', 'smallmol_glog_annotated_curated.rds'))

# Extract the assay matrix
smallmol_matrix <- as.matrix(data.frame(assay(smallmol_SE), row.names = rowData(smallmol_SE)$molecule_name))
saveRDS(smallmol_matrix, here('input', '05_MOFA', 'smallmol_matrix.rds'))
```