---
title: "Breathing Together Project"
subtitle: "Multiomic Analysis of BT groups 2-4"
author: "Matthew Macowan"
date: "`r format(Sys.time())`"
output:
  html_document:
  toc: true
number_sections: true
toc_float: true
toc_depth: 4
---

Copyright (c) 2023, Mucosal Immunology Lab, Monash University, Melbourne, Australia.

# Environment setup

## 1) Load packages

```{r load packages, eval=TRUE, collapse=TRUE, echo=FALSE}
# Get R version and OS information
R.version$version.string
R.version$platform

# Install packages requiring separate installation
library(devtools)
remotes::install_github("GraceYoon/SPRING")

# Load R packages
pkgs <- c('knitr', 'here', 'ggplot2', 'tidyverse', 'ggpubr', 'ggsci', 'ggbeeswarm', 'compositions',
          'rstatix', 'purrr', 'phyloseq', 'decontam', 'phylotools', 'microbiome', 'metagenomeSeq',
          'plotly', 'MOFA2', 'edgeR', 'SummarizedExperiment', 'WGCNA', 'caret', 'scales',
          'doParallel', 'clusterProfiler', 'org.Hs.eg.db', 'ReactomePA', 'KEGGREST', 'methods',
          'MultiAssayExperiment', 'reshape', 'ggpmisc', 'ComplexHeatmap', 'glmnet', 'hdi',
          'seriation', 'circlize', 'ggside', 'SNFtool', 'mogsa', 'CancerSubtypes', 'gdata',
          'DESeq2', 'matrixStats', 'factoextra', 'psych', 'ggcorrplot', 'kableExtra',
          'pathfindR', 'FactoMineR', 'factoextra', 'pathview', 'gtools', 'stabs',
          'ggrepel', 'strex', 'sva')

pacman::p_load(char = pkgs)

for (pkg in pkgs) {
  print(paste0(pkg, ': ', packageVersion(pkg)))
}

# Set seed
set.seed(2)

# Set colours for disease groups
BT_colours <- c('2_Preschool_control' = 'darkseagreen3',
                '3_Preschool_wheeze' = 'orange3',
                '4_School_asthma' = 'indianred',
                'Control' = 'darkseagreen3',
                'Wheeze_Asthma' = 'indianred',
                'Low' = '#015074',
                'Mid' = '#EFCD93',
                'High' = '#AC3631')

# Set colours for omics modalities
omic_colours <- c('bact_gk' = '#00a14b',
                  'host_rna' = '#29abe2',
                  'small_mol' = '#fbb03b')

# Load custom scripts
source(here::here('scripts', 'rotate_df.R'))
source(here::here('scripts', 'plot_list_to_pdf.R'))
source(here::here('scripts', 'bio_limma.R'))
```

# MOFA

## 1) Define MOFA code chunks

In order to make sure that the code is as readable as possible, we will define various code chunks here that define the steps for MOFA downstream.

### a) Prepare MultiAssayExperiment and check distributions

Firstly we need to be able to assemble our data into a MultiAssayExperiment object. We can then check the distribution and skew of the data.

```{r, eval=FALSE}
# Define the function
prepare_MAE <- function(omics_list, metadata, plot_subtitle) {
  omics_list <- omics_list
  # Ensure unique feature names with maximum length of 50 characters
  omics_name_mapper <- list()
  for (i in seq_along(omics_list)) {
    name_mapper <- data.frame('full_name' = as.character(rownames(omics_list[[i]]))) %>%
      remove_rownames() %>%
      mutate(name_short = str_sub(full_name, 1, 48)) %>%
      mutate(name_short = make.unique(name_short))
    omics_name_mapper[[i]] <- name_mapper
    rownames(omics_list[[i]]) <- name_mapper$name_short
  }
  # Create a MultiAssayExperiment object
  omics_mae <- MultiAssayExperiment(experiments = omics_list, colData = DataFrame(metadata))
  # Loop through datasets and inspect the distributions
  omics_dist_plots <- list()
  for (i in seq_along(omics_mae@ExperimentList)) {
    dataset_name <- names(omics_mae@ExperimentList)[i]
    data <- melt(omics_mae@ExperimentList[[i]])
    s <- skew(data$value)
    title = paste0(dataset_name, ' - skew: ', round(s, 3))
    p <- ggdensity(data, x = 'value', fill = 'lightgrey', title = title) +
      labs(subtitle = plot_subtitle) +
      stat_overlay_normal_density(color = 'red', linetype = 'dashed') +
      theme(text = element_text(size = 7))
    omics_dist_plots[[dataset_name]] <- p
  }
  omics_dist_plots_arr <- ggarrange(plotlist = omics_dist_plots, ncol = 1)
  # Combine the necessary items and return as a list
  return(list(omics_mae = omics_mae,
              omics_name_mapper = omics_name_mapper,
              omics_dist_plots = omics_dist_plots_arr))
}
```

### b) Running of MOFA and correction of factors

Now we need to prepare the code to run MOFA with the MultiAssayExperiment object we created above.

```{r, eval=FALSE}
# Define the function
run_MOFA <- function(omics_mae, drop_factor_threshold, output_file) {
  # Create the MOFA object from the MultiAssayExperiment
  MOFAobject <- create_mofa(omics_mae)
  MOFAobject <- set_covariates(MOFAobject, covariates = 'Age_months')
  # Generate data overview plot
  data_overview_plot <- plot_data_overview(MOFAobject)
  # Set MOFA parameters
  (DataOptions <- get_default_data_options(MOFAobject))
  DataOptions$center_groups <- TRUE
  DataOptions$scale_views <- TRUE
  DataOptions$scale_groups <- TRUE
  (ModelOptions <- get_default_model_options(MOFAobject))
  ModelOptions$num_factors <- 8
  ModelOptions$spikeslab_factors <- FALSE
  ModelOptions$spikeslab_weights <- FALSE
  (TrainOptions <- get_default_training_options(MOFAobject))
  TrainOptions$maxiter <- 10000 # 10000
  TrainOptions$drop_factor_threshold <- drop_factor_threshold
  TrainOptions$seed <- 2
  TrainOptions$convergence_mode <- 'slow'
  (MefistoOptions <- get_default_mefisto_options(MOFAobject))
  # Prepare the MOFA object
  MOFAobject <- prepare_mofa(MOFAobject,
                             data_options = DataOptions, 
                             model_options = ModelOptions, 
                             training_options = TrainOptions,
                             mefisto_options = MefistoOptions)
  # Run MOFA
  MOFArun <- run_mofa(MOFAobject, outfile = here::here('output', '05_MOFA', '01_MOFArun', 'tmp', 'MOFArun_tmp'), use_basilisk = TRUE)
  saveRDS(MOFArun, output_file)
  # Return the MOFA run output
  return(MOFArun)
}
```

### c) Save functions to RDS

```{r, eval=FALSE}
# Add functions to a list and save to RDS
custom_MOFA_functions <- list(prepare_MAE = prepare_MAE,
                              run_MOFA = run_MOFA)
saveRDS(custom_MOFA_functions, here('input', '05_MOFA', 'custom_MOFA_functions.rds'))
```

## 2) Run MOFA

Using the code chunks above, we will now run MOFA on our different datasets.

### a) Recover datasets

```{r, eval=FALSE}
# Recover the different datasets
host_rna_matrix <- readRDS(here('input', '05_MOFA', 'host_rna_bronch_matrix_min100.rds'))
metaG_CLR_matrix <- readRDS(here('input', '05_MOFA', 'metaG_CLR_matrix.rds'))
small_mol_matrix <- readRDS(here('input', '05_MOFA', 'smallmol_matrix.rds'))

# Import the metadata
metadata_mofa <- readRDS(here('input', '00_Metadata', 'metadata_mofa.rds')) %>%
  column_to_rownames(var = 'Subject_ID')

# Recover the functions from above
custom_MOFA_functions <- readRDS(here('input', '06_MOFA', '01_TestingStrategy', 'custom_MOFA_functions.rds'))

# Add the custom functions to the environment using the base R assign function
for (i in seq_along(custom_MOFA_functions)) {
  assign(names(custom_MOFA_functions)[i], custom_MOFA_functions[[i]])
}
```

### b) Prepare MultiAssayExperiments

We will now prepare a MultiAssayExperiment to combine our datasets into a single object.

```{r, eval=FALSE}
# Describe the input data to use as a subtitle for the data distribution plots
test_name <- paste0('hostRNA=min100; ', 
                    'bactfilt=d1p0.1; ',
                    'small_mol=full; ',
                    'factordrop=0.05')

# Create a list with the necessary datasets
omics_list <- list(host_rna = host_rna_matrix,
                   bact_gk = metaG_CLR_matrix,
                   small_mol = small_mol_matrix)

# Use the prepare_MAE function to prepare the MultiAssayExperiment object
mofa_mae <- prepare_MAE(omics_list = omics_list,
                        metadata = metadata_mofa,
                        plot_subtitle = test_name)

# Inspect the distribution plots
mofa_mae$omics_dist_plots
```

### c) Run MOFA

Now we have our MultiAssayExperiment, we can actually run MOFA+.

```{r, eval=FALSE}
# Set file output name for MOFA object
mofa_output_filename <- here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))

# Run MOFA data integration on the MultiAssayExperiment
MOFA <- run_MOFA(omics_mae = mofa_mae$omics_mae,
                 drop_factor_threshold = 0.05,
                 output_file = mofa_output_filename)
```

## 3) Overview of MOFA output

### a) Variance explained

```{r}
# Recover MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))

# Extract variance explained data
var_exp <- data.frame(MOFA@cache$variance_explained$r2_per_factor$group1) %>%
  rownames_to_column(var = 'Factor') %>%
  melt() %>%
  filter(value > 0.01)

# Plot variance explained by factor
(var_exp_plot <- ggplot(var_exp, aes(x = Factor, y = value)) +
    geom_col(aes(fill = variable), position = 'stack') +
    scale_fill_manual(values = omic_colours, name = 'Omic') +
    geom_text(aes(label = round(value, 2), group = variable), position = position_stack(vjust = 0.5)) +
    labs(title = 'Variance explained by MOFA+ factor',
         y = 'Variance explained (%)') +
    theme_pubr(legend = 'right'))
ggsave(here('figures', '05_MOFA', '01_Overview', 'var_explained_plot.pdf'),
       var_exp_plot, width = 14, height = 12, units = 'cm')
```

### b) Determine factors associated with sample group and disease progression

```{r}
# Recover MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))
MOFA_impute <- impute(MOFA)

# Get factor data
factor_data <- data.frame(MOFA2::get_factors(MOFA_impute)$group1) %>%
    rownames_to_column(var = 'Subject_ID') %>%
    left_join(data.frame(MOFA@samples_metadata, row.names = samples_names(MOFA)$group1) %>%
                rownames_to_column(var = 'Subject_ID'),
              by = 'Subject_ID') %>%
    column_to_rownames(var = 'Subject_ID')

# Define function to test differences in MOFA factors according to a categorical covariate
factor_wilcox_test <- function(factor_data, covariate) {
  factor_colnames <- colnames(factor_data %>% dplyr::select(contains('Factor')))
  wilcox_list <- list()
  for (i in factor_colnames) {
    w <- wilcox_test(data = factor_data, formula = as.formula(paste0(i, ' ~ ', covariate)))
    wilcox_list[[i]] <- w
  }
  wilcox_list <- do.call(rbind, wilcox_list)
}

# BT group
bt_group_wilcox <- factor_wilcox_test(factor_data, 'BT_group') %>%
    arrange(p.adj) %>%
    mutate(comparison = paste0(group1, ' vs. ', group2)) %>% 
    dplyr::select(-group1, -group2) %>%
    dplyr::select(`.y.`, comparison, everything()) %>%
    group_by(comparison) %>%
    slice_head(n = 2) %>%
    ungroup() %>%
    filter(p.adj < 0.05) %>%
    arrange(p.adj)
kable(bt_group_wilcox) %>%
  kable_styling(bootstrap_options = 'striped')
```

Latent factor 4 shows the most significant differences between BT groups, so we will investigate this further.

## 4) Characterisation of Latent Factor 4

Factor 4 captures some of the differences in the BT group, showing significant differences between all three groups. 
We should determine what features drive variation in this factor.

### a) Overview

Firstly we will create a boxplot, density plot, and check differences between atopic vs. non-atopic samples.

```{r}
# Recover MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))
MOFA_impute <- impute(MOFA)

# Get factor data
factor_data <- data.frame(MOFA2::get_factors(MOFA_impute)$group1) %>%
  rownames_to_column(var = 'Subject_ID') %>%
  left_join(data.frame(MOFA@samples_metadata, row.names = samples_names(MOFA)$group1) %>%
              rownames_to_column(var = 'Subject_ID'),
            by = 'Subject_ID') %>%
  left_join(read_csv(here('input', '00_Metadata', 'metadata_extra.csv'), col_types = 'cnfffffffffnffnfn'),
            by = c('Subject_ID', 'Gest_age_birth')) %>%
  column_to_rownames(var = 'Subject_ID')
factor_data[factor_data == 'nan'] <- NA
saveRDS(factor_data, here('input', '00_Metadata', 'factor_data_extra.rds'))
write_csv(factor_data, here('input', '00_Metadata', 'factor_data_extra.csv'))

# Plot Factor 4 vs. BT group
(f4_BTgroup_plot <- ggplot(factor_data, aes(x = BT_group, y = Factor4)) +
    geom_boxplot(aes(fill = BT_group), alpha = 0.6) +
    geom_dotplot(binaxis = 'y', stackdir = 'center') +
    scale_fill_manual(values = BT_colours) +
    guides(fill = 'none') +
    stat_compare_means(comparisons = list(c(1,2), c(2,3), c(1,3)), size = 3) +
    theme(text = element_text(size = 8)) +
    labs(title = 'Factor 4 - BT group') +
    theme_pubr(base_size = 8))
ggsave(here('figures', '05_MOFA', '02_Factor4', 'F4_BTgroup_boxplot.pdf'),
       f4_BTgroup_plot, width = 10.5, height = 16.5, units = 'cm')

# Make a density plot for Factor 4 coloured by BT group
(f4_density_plot <- ggplot(factor_data, aes(x = Factor4)) +
   geom_density(aes(fill = BT_group), alpha = 0.5) +
   scale_fill_manual(values = BT_colours) +
   labs(title = 'Density plot of BT groups along Factor 4',
        y = 'Density') +
   theme_pubr())
ggsave(here('figures', '05_MOFA', '02_Factor4', 'densityplot_BTgroup_factor4.pdf'),
       f4_density_plot, width = 18, height = 6, units = 'cm')

# Check measures of atopy
# 30IU/mL IgE threshold based on prescribing information for omalizumab as previously described (PMID: 31999898)
(atopy_f4 <- ggplot(atopy_data %>% filter(!is.na(atopy)), aes(x = atopy, y = Factor4)) +
    geom_boxplot(fill = 'grey90') +
    geom_dotplot(aes(fill = BT_group), binaxis = 'y', stackdir = 'center', dotsize = 0.5) +
    stat_compare_means(comparisons = list(c(1,2))) +
    scale_fill_manual(values = BT_colours) +
    labs(title = 'Atopy vs. Factor 4') +
    theme_pubr(x.text.angle = 45)
)
ggsave(here('figures', '05_MOFA', '02_Factor4', 'F4_atopy_plot.pdf'), atopy_f4,
       width = 12, height = 12, units = 'cm')
```

### b) Differential testing - bio_limma

#### i) HostRNA

```{r}
# Recover MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))
MOFA_impute <- impute(MOFA)

# Get factor data
factor_data <- data.frame(MOFA2::get_factors(MOFA_impute)$group1) %>%
  rownames_to_column(var = 'Subject_ID') %>%
  left_join(data.frame(MOFA@samples_metadata, row.names = samples_names(MOFA)$group1) %>%
              rownames_to_column(var = 'Subject_ID'),
            by = 'Subject_ID') %>%
  column_to_rownames(var = 'Subject_ID') %>%
  mutate(FactorFour = Factor4)

# Prepare datasets for limma
rna_dataset <- SummarizedExperiment(assays = MOFA_impute@imputed_data$host_rna$group1)
rna_dataset@metadata$metadata <- factor_data
rowData(rna_dataset)$shortname <- rownames(rna_dataset)

# Run bio_limma
rna_f4_limma <- bio_limma(input_data = rna_dataset,
                          model_formula_as_string = '~ FactorFour + BT_group + Age_months',
                          coefficients = 2,
                          adj_pval_threshold = 0.05,
                          logFC_threshold = 0.5,
                          redo_boxplot_stats = TRUE,
                          use_groups_as_subtitle = TRUE,
                          force_feature_table_variable = 'BT_group',
                          plot_output_folder = here('figures', '05_MOFA', '02_Factor4', 'hostRNA', 'limmaDA'),
                          plot_file_prefix = 'mofa_f4')
saveRDS(rna_f4_limma, here('figures', '05_MOFA', '02_Factor4', 'hostRNA', 'limmaDA', 'rna_f4_biolimma.rds'))
```

```{r}
# Plot new barplot (instead of the one provided by the bio_limma function)
rna_f4_limma <- readRDS(here('figures', '05_MOFA', '02_Factor4', 'hostRNA', 'limmaDA', 'rna_f4_biolimma.rds'))
rna_barplot_df <- rna_f4_limma$limma_significant$FactorFour %>%
  rownames_to_column(var = 'Gene') %>%
  mutate(sig_colour = -log10(adj.P.Val),
         direction = ifelse(logFC > 0, 'Disease', 'Health')) %>%
  group_by(direction) %>%
  arrange(desc(abs(logFC))) %>%
  slice_head(n = 15) %>%
  ungroup() %>%
  arrange(logFC) %>%
  mutate(Gene = factor(Gene, levels = Gene))

(rna_f4_barplot <- ggplot(rna_barplot_df, aes(x = logFC, y = Gene)) +
    geom_col(aes(fill = direction)) +
    scale_fill_npg(name = '-log10(adj. p-value', alpha = 0.8) +
    labs(title = 'Top Factor-characterising Genes') +
    theme_pubr())
ggsave(here('figures', '05_MOFA', '02_Factor4', 'hostRNA', 'limmaDA', 'top_f4_genes_barplot.pdf'),
       rna_f4_barplot, width = 20, height = 16, units = 'cm')
```

### 2) Bacterial GhostKOALA data

```{r}
# Recover MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))
MOFA_impute <- impute(MOFA)

# Get factor data
factor_data <- data.frame(MOFA2::get_factors(MOFA_impute)$group1) %>%
  rownames_to_column(var = 'Subject_ID') %>%
  left_join(data.frame(MOFA@samples_metadata, row.names = samples_names(MOFA)$group1) %>%
              rownames_to_column(var = 'Subject_ID'),
            by = 'Subject_ID') %>%
  column_to_rownames(var = 'Subject_ID') %>%
  mutate(FactorFour = Factor4)

# Prepare datasets for limma
gk_dataset <- SummarizedExperiment(assays = MOFA_impute@imputed_data$bact_gk$group1)
gk_dataset@metadata$metadata <- factor_data
rowData(gk_dataset)$shortname <- rownames(gk_dataset)

# Run bio_limma
gk_f4_limma <- bio_limma(input_data = gk_dataset,
                         model_formula_as_string = '~ FactorFour + BT_group + Age_months',
                         coefficients = 2,
                         adj_pval_threshold = 0.05,
                         logFC_threshold = 0.25,
                         redo_boxplot_stats = TRUE,
                         use_groups_as_subtitle = TRUE, 
                         feature_plot_dot_colours = 'BT_group',
                         plot_output_folder = here('figures', '05_MOFA', '02_Factor4', 'bactGhostKOALA', 'limmaDA'),
                         plot_file_prefix = 'mofa_f4')
saveRDS(gk_f4_limma, here('figures', '05_MOFA', '02_Factor4', 'bactGhostKOALA', 'limmaDA', 'bactgk_f4_biolimma.rds'))
```

```{r}
# Recover significant features
gk_f4_limma <- readRDS(here('figures', '05_MOFA', '02_Factor4', 'bactGhostKOALA', 'limmaDA', 'bactgk_f4_biolimma.rds'))
bactgk_f4_limma_sig <- gk_f4_limma$limma_significant$FactorFour

# Prepare summarised data for generation of a new barplot of top features
bact_barplot_df <- bactgk_f4_limma_sig %>%
  rownames_to_column(var = 'Feature') %>%
  mutate(Species = gsub('_.*', '', Feature),
         direction = ifelse(logFC > 0, 'Disease', 'Health')) %>%
  group_by(Species, direction) %>%
  summarise(count = n()) %>% # count the number of DA genes per taxa
  ungroup() %>%
  group_by(Species) %>%
  summarise(direction = direction, count = count, bact_total = sum(count)) %>% # summarise the total counts up and down along Factor 4
  ungroup() %>%
  filter(bact_total > 10 & Species != 'Unknown') %>% # Remove rows where the taxonomy is unknown
  mutate(count = ifelse(direction == 'Disease', count, -count)) %>% # Fix count direction for use along the x-axis of a barplot
  group_by(Species) %>%
  mutate(count_diff = sum(count)) %>%
  mutate(logcount_diff = log(abs(count_diff))) %>% # normalise summed DA gene counts by log-normalisation (some are very large)
  mutate(logcount_diff = ifelse(count_diff > 0, logcount_diff, -logcount_diff),
         logcount = ifelse(count > 0, log(abs(count)), -log(abs(count)))) %>%
  mutate(logcount_midpoint = mean(logcount)) %>% # find the midpoint of changes to determine overall direction of change (mean of 2 values = midpoint)
  ungroup() %>%
  arrange(logcount_midpoint) %>% # sort the shifts from highest in health to highest in disease
  mutate(Species = factor(Species, levels = unique(Species))) %>%
  dplyr::select(Species, logcount_midpoint) %>%
  distinct() %>%
  mutate(direction = ifelse(logcount_midpoint > 0, 'Disease', 'Health')) %>%
  arrange(logcount_midpoint) %>%
  mutate(Species = factor(Species, levels = Species))

# Plot new barplot (instead of the one provided by the bio_limma function)
(gk_f4_taxplot <- ggplot(bact_barplot_df, aes(x = logcount_midpoint, y = Species)) +
    geom_col(aes(fill = direction)) +
    scale_fill_npg(name = 'Direction', alpha = 0.8) +
    labs(title = 'Factor-associated changes in bacterial genes',
         subtitle = 'Taxa with >10 DE genes (FDR < 0.05; logFC > 0.25)',
         x = 'log(DE gene count)',
         y = 'Taxa') +
    theme_pubr())
ggsave(here('figures', '05_MOFA', '02_Factor4', 'bactGhostKOALA', 'limmaDA', 'gk_f4_taxa_barplot.pdf'),
       gk_f4_taxplot, width = 20, height = 14, units = 'cm')

# Save as an RDS for addition of the transcriptional activity plot (see '03_Metatranscriptomics.Rmd')
saveRDS(gk_f4_taxplot, here('figures', '05_MOFA', '02_Factor4', 'bactGhostKOALA', 'limmaDA', 'gk_f4_taxa_barplot.rds'))
```



## 4) Characterisation of Latent Factor 2

Factor 2 captures a lot of variation explained by small molecules, so we should also investigate what this factor explains.

### a) Overview

Firstly we will create a boxplot, density plot, and check differences between atopic vs. non-atopic samples.

```{r}
# Recover selected MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))
MOFA_impute <- impute(MOFA)

# Recover factor data
factor_data <- readRDS(here('input', '00_Metadata', 'factor_data_extra.rds'))

# Plot Factor 2 vs. BT group
(f2_BTgroup_plot <- ggplot(factor_data, aes(x = BT_group, y = Factor2)) +
    geom_boxplot(aes(fill = BT_group), alpha = 0.6) +
    geom_dotplot(binaxis = 'y', stackdir = 'center') +
    scale_fill_manual(values = BT_colours) +
    guides(fill = 'none') +
    stat_compare_means(comparisons = list(c(1,2), c(2,3), c(1,3)), size = 3) +
    theme(text = element_text(size = 8)) +
    labs(title = 'Factor 2 - BT group') +
    theme_pubr(base_size = 8))
ggsave(here('figures', '05_MOFA', '03_Factor2', 'F2_BTgroup_boxplot.pdf'),
       f2_BTgroup_plot, width = 10.5, height = 16.5, units = 'cm')

# Make a density plot for Factor 2 coloured by BT group
(f2_density_plot <- ggplot(factor_data, aes(x = Factor2)) +
   geom_density(aes(fill = BT_group), alpha = 0.5) +
   scale_fill_manual(values = BT_colours) +
   labs(title = 'Density plot of BT groups along Factor 2',
        y = 'Density') +
   theme_pubr())
ggsave(here('figures', '05_MOFA', '03_Factor2', 'densityplot_BTgroup_factor2.pdf'),
       f2_density_plot, width = 18, height = 6, units = 'cm')

# Check measures of atopy
# 30IU/mL IgE threshold based on prescribing information for omalizumab as previously described (PMID: 31999898)
(atopy_f2 <- ggplot(atopy_data %>% filter(!is.na(atopy)), aes(x = atopy, y = Factor2)) +
    geom_boxplot(fill = 'grey90') +
    geom_dotplot(aes(fill = BT_group), binaxis = 'y', stackdir = 'center', dotsize = 0.5) +
    stat_compare_means(comparisons = list(c(1,2))) +
    scale_fill_manual(values = BT_colours) +
    labs(title = 'Atopy vs. Factor 2') +
    theme_pubr(x.text.angle = 45)
)
ggsave(here('figures', '05_MOFA', '03_Factor2', 'F2_atopy_plot.pdf'), atopy_f2,
       width = 12, height = 12, units = 'cm')
```

### b) Differential testing - bio_limma

##### 1) HostRNA

```{r}
# Recover MOFA
MOFA <- readRDS(here('output', '05_MOFA', '01_MOFArun', 'MOFA_output.rds'))
MOFA_impute <- impute(MOFA)

# Get factor data
factor_data <- data.frame(MOFA2::get_factors(MOFA_impute)$group1) %>%
  rownames_to_column(var = 'Subject_ID') %>%
  left_join(data.frame(MOFA@samples_metadata, row.names = samples_names(MOFA)$group1) %>%
              rownames_to_column(var = 'Subject_ID'),
            by = 'Subject_ID') %>%
  column_to_rownames(var = 'Subject_ID') %>%
  mutate(FactorTwo = Factor2)

# Prepare datasets for limma
rna_dataset <- SummarizedExperiment(assays = MOFA_impute@imputed_data$host_rna$group1)
rna_dataset@metadata$metadata <- factor_data
rowData(rna_dataset)$shortname <- rownames(rna_dataset)

# Run bio_limma
rna_f2_limma <- bio_limma(input_data = rna_dataset,
                          model_formula_as_string = '~ FactorTwo + BT_group + Age_months',
                          coefficients = 2,
                          adj_pval_threshold = 0.05,
                          logFC_threshold = 0.5,
                          redo_boxplot_stats = TRUE,
                          use_groups_as_subtitle = TRUE,
                          force_feature_table_variable = 'BT_group',
                          plot_output_folder = here('figures', '05_MOFA', '03_Factor2', 'hostRNA', 'limmaDA'),
                          plot_file_prefix = 'mofa_f2')
saveRDS(rna_f2_limma, here('figures', '05_MOFA', '03_Factor2', 'hostRNA', 'limmaDA', 'rna_f2_biolimma.rds'))

# Plot new barplot (instead of the one provided by the bio_limma function)
rna_f2_limma <- readRDS(here('figures', '05_MOFA', '03_Factor2', 'hostRNA', 'limmaDA', 'rna_f2_biolimma.rds'))
rna_barplot_df <- rna_f2_limma$limma_significant$FactorTwo %>%
  rownames_to_column(var = 'Gene') %>%
  mutate(sig_colour = -log10(adj.P.Val),
         direction = ifelse(logFC > 0, 'Wheeze', 'Control_Asthma')) %>%
  group_by(direction) %>%
  arrange(desc(abs(logFC))) %>%
  slice_head(n = 15) %>%
  ungroup() %>%
  arrange(logFC) %>%
  mutate(Gene = factor(Gene, levels = Gene))

(rna_f2_barplot <- ggplot(rna_barplot_df, aes(x = logFC, y = Gene)) +
    geom_col(aes(fill = direction)) +
    scale_fill_npg(name = '-log10(adj. p-value', alpha = 0.8) +
    labs(title = 'Top Factor-characterising Genes') +
    theme_pubr())
ggsave(here('figures', '05_MOFA', '03_Factor4', 'hostRNA', 'limmaDA', 'top_f2_genes_barplot.pdf'),
       rna_f4_barplot, width = 20, height = 16, units = 'cm')
```